---
title: "mediationverse"
subtitle: "A Unified Ecosystem for Causal Mediation Analysis in R"
---

```{r}
#| echo: false
#| message: false
# Load package for examples
library(mediationverse)
```

## Overview

**mediationverse** is a meta-package that provides a unified interface to a comprehensive ecosystem of R packages for causal mediation analysis. With a single `library(mediationverse)` call, you get access to state-of-the-art methods for:

- **Model fitting** via [medfit](https://data-wise.github.io/medfit/)
- **Effect size calculation** via [probmed](https://data-wise.github.io/probmed/)
- **Confidence intervals** via [RMediation](https://data-wise.github.io/rmediation/)
- **Sensitivity analysis** via [medrobust](https://data-wise.github.io/medrobust/)
- **Simulation studies** via [medsim](https://data-wise.github.io/medsim/)

## Installation

```r
# Install from GitHub
# install.packages("pak")
pak::pak("Data-Wise/mediationverse")
```
## Quick Start

```{r}
#| eval: false
# Load the ecosystem
library(mediationverse)

# Check which packages are available
mediationverse_packages()

# Update all packages
mediationverse_update()
```

## Ecosystem Status

```{r}
#| echo: false
mediationverse_packages()
```

## Learn More

- [Getting Started](vignettes/getting-started.qmd) - Installation and basic usage
- [Workflow Guide](vignettes/mediationverse-workflow.qmd) - Complete analysis workflow
- [Function Reference](reference/index.qmd) - API documentation

## Citation

If you use mediationverse in your research, please cite:

```
@software{mediationverse,
  author = {Tofighi, Davood},
  title = {mediationverse: A Unified Ecosystem for Causal Mediation Analysis in R},
  year = {2025},
  url = {https://github.com/Data-Wise/mediationverse}
}
```
